Spatiotemporal Multi-Resolution Approximation of the Amari Type Neural Field Model

### 2011-11-08 Multiresolution_framework/ltx/manuscript.tex
for merge
added a section on multires neural fields

Neural field models describe the mass action of the central nervous system and are a critical link in our understanding of the biophysics of the EEG. These models typically describe the macroscopic dynamics of the human brain, but are also descriptive of finer-scale neurodynamics. The ability to create patient-specific neural models will contribute to our knowledge base of diseases such as epilepsy and will enable the development of new treatment strategies. This is particularly relevant to the advent of new devices that use therapeutic electrical stimulation. Stimulation strategies for devices currently operate in an open loop, where stimulation parameters are chosen using a process of trial and error. Therefore, there exists enormous potential to improve the performance of these devices using systems theory, which in turn requires a suitable dynamic model. Neural mass and neural field models are ideal candidates for estimation algorithms due to their strong links with physiology and their parsimony. It is expected that parameters of the neural fields models will be patient-specific and will, therefore, need to be inferred from data.

The first work describing data-driven mesoscopic neural modeling used a neural mass model to fit EEG data~\cite{Valdes1999}. This approach was extended to coupled neural masses through a Bayesian estimation scheme dubbed Dynamic Causal Modeling (DCM)~\cite{David2003}. Following this work, data-driven modeling was extended to continuum field equations that explained the richer dynamics of spatiotemporal neural fields \cite{Galka2008,schiff2008kalman,Daunizeau2009}. Most recently, a framework was developed where a finite element model (FEM) of the neural field (via a global Galerkin projection) was formed, using a basis function decomposition, to transform the PDE neural field equations into a finite dimension system to facilitate efficient state and parameter estimation~\cite{Freestone2011}. The current paper is an extension to this recent study, where we derive a neural field state-space model that accounts for the multi-resolution architecture and spatial dynamics of the human cortex. In this way, a flexible framework is created, whereby both macroscopic and microscopic behavior of the system can be represented simultaneously.

@@

Equations \eqref{eq:KernelExpansion} and \eqref{eq:FieldExpansion} are infinite series expansions and must be truncated at some level $j$. The field and the connectivity kernel are approximated by
\begin{align}
	w\left(r-r'\right) &\approx \boldsymbol\theta^\top\boldsymbol\lambda\left(r-r'\right) 
	\label{eq:KernelFiniteExpansion} \\
	v_t\left(r\right) &\approx \boldsymbol\mu^\top\left(r\right)\mathbf{x}_t,
	\label{eq:FieldFiniteExpansion}
\end{align}
where the unknown parameter and state vectors, $\boldsymbol\theta \in \mathbb{R}^{n_{\theta}}$ and $\mathbf{x}_t \in \mathbb{R}^{n_x}$, are defined as 
\begin{align}
\boldsymbol\theta^\top &=[\begin{array}{ccccc} \boldsymbol\alpha_{j_0}^\top & \boldsymbol\beta_{j_0}^\top & \boldsymbol\beta_{j_0+1}^\top & \cdots & \boldsymbol\beta_{j}^\top \end{array}] 
\label{KernelWeights} \\
\mathbf{x}_{t}^\top &=[\begin{array}{ccccc}\mathbf{x}_{t,j_{0}}^\top &  \check{\mathbf{x}}_{t,j_{0}}^\top & \check{\mathbf{x}}_{t,j_{0}+1}^\top & \cdots & \check{\mathbf{x}}_{t,j}^\top\end{array}].
\label{FieldWeights}
\end{align}
The kernel and field approximation coefficient vectors, $\boldsymbol \alpha_{j_0}$ and $\mathbf{x}_{t,j_{0}}$, contain all the coefficients $\left\lbrace\alpha_{j_0, l}:l \in \mathbb{Z} \right\rbrace $ and $\left\lbrace x_{t,j_0, l}: l \in \mathbb{Z}\right\rbrace$, respectively. Similarly, the kernel and the field detail coefficient vectors, $\boldsymbol\beta_{j}$ and $\check{\mathbf{x}}_{t,j}$, contain all the coefficients $\left\lbrace \beta_{j,l} :l \in \mathbb{Z}\right\rbrace$ and $\left\lbrace  \check x_{t,j, l}:l \in \mathbb{Z}\right\rbrace$, respectively.

>>>>

Neural field models describe the mass action of the central nervous system and are a critical link in our understanding of the biophysics of the EEG. These models typically describe the macroscopic dynamics of the human brain, but are also descriptive of finer-scale neurodynamics. For example.. integrate and fire... The utility in these models to answer questions in neuroscience lies in the ability to infer parameters from data. The ability to create patient-specific neural models will contribute to our knowledge base of diseases such as epilepsy and will enable the development of new treatment strategies. This is particularly relevant to the advent of new devices that use therapeutic electrical stimulation. Stimulation strategies for devices currently operate in an open loop, where stimulation parameters are chosen using a process of trial and error. Therefore, there exists enormous potential to improve the performance of these devices using systems theory, which in turn requires a suitable dynamic model. Neural mass and neural field models are ideal candidates for estimation algorithms due to their strong links with physiology and their parsimony. It is expected that parameters of the neural fields models will be patient-specific and will, therefore, need to be inferred from data.

The first work describing data-driven mesoscopic neural modeling used a neural mass model to fit EEG data~\cite{Valdes1999}. This approach was extended to coupled neural masses through a Bayesian estimation scheme dubbed Dynamic Causal Modeling (DCM)~\cite{David2003}. Following this work, data-driven modeling was extended to continuum field equations that explained the richer dynamics of spatiotemporal neural fields \cite{Galka2008,schiff2008kalman,Daunizeau2009}. Most recently, a framework was developed where a finite element model (FEM) of the neural field (via a global Galerkin projection) was formed, using a basis function decomposition, to transform the PDE neural field equations into a finite dimension system to facilitate efficient state and parameter estimation~\cite{Freestone2011}. This paper is an extension to the aforementioned study, where we derive a neural field state-space model that accounts for the multi-resolution architecture and spatial dynamics of the human cortex. In this way, a flexible framework is created, whereby both macroscopic and microscopic behavior of the system can be represented simultaneously \dean{and completely}.

\section{Multi-Resolution Cortical Architecture}
The human neocortex has an inherent multi-resolution architecture. On finer scales... where any measurable spatiotemporal neural field can be written as 
\begin{equation}
	v\left(\mathbf{r},t\right) = \sum_{j} v_j\left(\mathbf{r},t\right),
\end{equation}
where $v$ is a voltage field that may be measured with a multi-electrode array or voltage sensitive dye for example, $\mathbf{r}\in \mathbb{R}^n$ (where $n\le3$) describes the spatial location with the field, $t$ denotes time, and $j$ indexes the different levels of resolution in the field.

This multi-resolution structure has been described by numerous methods (Wilson and Cowan). This idea of multi-resolution can be thought of a generalisation of modelling the layers of the cortex separately, where the connectivity kernel, or neural foot-print, has varying widths.

Typically, the experimentalist only has access the net field generated by the multi-resolution architecture. Follow this, it is natural to ask how do we best describe the multi-resolution structure of the neural system. The solution to this problem was proposed by \dean{Breakspear and Stam} in using a wavelet decomposition.  
\begin{figure}[t]
	\centering
		\includegraphics[scale=0.65]{./Graph/MultiresNeuralField.pdf}
	\caption{Multi-layer neural field model.}
	\label{fig:MultiLayerFieldModel}
\end{figure}

@@

Equations \ref{eq:KernelExpansion} and \ref{eq:FieldExpansion} are infinite series expansions and must be truncated at some level $j$. The field and the connectivity kernel are approximated by
\begin{align}
	w\left(r-r'\right) &\approx \boldsymbol\theta^\top\boldsymbol\lambda\left(r-r'\right) 
	\label{eq:KernelFiniteExpansion} \\
	v_t\left(r\right) &\approx \boldsymbol\mu^\top\left(r\right)\mathbf{x}_t,
	\label{eq:FieldFiniteExpansion}
\end{align}
where the unknown parameter and state vectors, $\boldsymbol\theta \in \mathbb{R}^{n_{\theta}}$ and $\mathbf{x}_t \in \mathbb{R}^{n_x}$, are defined as 
\begin{align}
\boldsymbol\theta^\top &=[\begin{array}{ccccc} \boldsymbol\alpha_{j_0}^\top & \boldsymbol\beta_{j_0}^\top & \boldsymbol\beta_{j_0+1}^\top & \cdots & \boldsymbol\beta_{j}^\top \end{array}] 
\label{KernelWeights} \\
\mathbf{x}_{t}^\top &=[\begin{array}{ccccc}\mathbf{x}_{t,j_{0}}^\top &  \check{\mathbf{x}}_{t,j_{0}}^\top & \check{\mathbf{x}}_{t,j_{0}+1}^\top & \cdots & \check{\mathbf{x}}_{t,j}^\top\end{array}].
\label{FieldWeights}
\end{align}
The kernel and field approximation coefficient vectors, $\boldsymbol \alpha_{j_0}$ and $\mathbf{x}_{t,j_{0}}$, contain all the coefficients $\left\lbrace\alpha_{j_0, l}:l \in \mathbb{Z} \right\rbrace $ and $\left\lbrace x_{t,j_0, l}: l \in \mathbb{Z}\right\rbrace$, respectively. Similarly, the kernel and the field detail coefficient vectors, $\boldsymbol\beta_{j}$ and $\check{\mathbf{x}}_{t,j}$, contain all the coefficients $\left\lbrace \beta_{j,l} :l \in \mathbb{Z}\right\rbrace$ and $\left\lbrace  \check x_{t,j, l}:l \in \mathbb{Z}\right\rbrace$, respectively.
